---
title: "Homework 13"
format: html
---

# 4.15.26
```{r}
library(tidyverse)
library(lubridate)
library(lme4)
library(car)
library(ggplot2)

bee_dat <- read_csv("Burnham_field_data_pathogens_wide.csv")

bee_dat <- bee_dat %>%
  mutate(
    sampling_date = mdy(sampling_date),
    site_code = factor(site_code),
    field_id = factor(field_id),
    bee_caste = factor(bee_caste),
    bombus_spp = factor(bombus_spp),
    host_plant = factor(host_plant),
    sampling_event = factor(sampling_event),
    sampling_event_num = as.numeric(as.character(sampling_event)),
    log10_BQCV_load = log10(BQCV_pathogen_load + 1),
    log10_DWV_load = log10(DWV_pathogen_load + 1),
    log10_Nosema_load = log10(Nosema_pathogen_load + 1)
  )
glimpse(bee_dat)

df_filtered <- bee_dat[bee_dat$log10_DWV_load > 0 & bee_dat$log10_BQCV_load > 0, ]
```


# Question 1
### I used the Burnham_field_data to look at if different sampling events have an impact on disease abundance.
### This was a non-gaussian (binomial function) and used an interaction term (do sampling events affect disease abundance)

```{r}
library(dplyr)

bee_dat_q1 <- bee_dat %>%
  select(sampling_event,
    BQCV_pathogen_binary,
    DWV_pathogen_binary,
    Nosema_pathogen_binary) %>%
  pivot_longer(
    cols = c(BQCV_pathogen_binary, DWV_pathogen_binary, Nosema_pathogen_binary),
    names_to = "disease",
    values_to = "abundance"
  )

q1_plot <- ggplot(bee_dat_q1, aes(x = sampling_event, y = abundance, color = disease))+
  geom_smooth(method = "glm", method.args = list(family = binomial))+
  labs(
    title = "Disease Abundance Across Different Sampling Events",
    y = "Disease Abundance",
    x = "Sampling Event",
    color = "Disease"
  ) +
  geom_jitter()

plot(q1_plot)



bee_dat_stat <- glm(abundance ~ sampling_event, data = bee_dat_q1)
Anova(bee_dat_stat)
```


### Sampling events were catagorical, and when sampling you either have or don't have the disease, so I used this plot to show a binary relationship if a disease was present during the sampling event
### The p-value was 0.02134, which is not statistially significant, so sampling events don't affect disease abundance. I expected the data to not be statistically significant, as I didn't think the sampling events occured over a long period of time. 
### However, if the sampling event occured over a longer time period, then it may have been interesting to see if disease abundance was changing over time, but I don't know the time frame of sampling events

# Question 2 
# I looked at removing sampling events (random effects) to see if the type of host plants affects disease prevelence

```{r}
q2_dwv <- lmer(
  log10_DWV_load ~ host_plant + (1 | sampling_event), data = bee_dat
)

q2_bqcv <- lmer(
  log10_BQCV_load ~ host_plant + (1 | sampling_event), data = bee_dat
)

q2_nos <- lmer(
  log10_Nosema_load ~ host_plant + (1 | sampling_event), data = bee_dat
)


q2_plot_nos <- ggplot(bee_dat, aes(x = host_plant, y = Nosema_pathogen_load))+
  geom_point(color = "darksalmon")+
  labs(
    title = "Nosema Pathogen Abundance Across Different Host Plane Species",
    y = "Disease Abundance",
    x = "Host Plant",
    color = "Disease"
  )

q2_plot_nos + coord_flip()


q2_plot_bqcv <- ggplot(bee_dat, aes(x = host_plant, y = log10_BQCV_load))+
  geom_point(color = "goldenrod") +
  labs(
    title = "BQCV Pathogen Abundance Across Different Host Plane Species",
    y = "Disease Abundance",
    x = "Host Plant",
    color = "Disease"
  )
q2_plot_bqcv + coord_flip()



q2_plot_dwv <- ggplot(bee_dat, aes(x = host_plant, y = log10_DWV_load))+
  geom_point(color = "royalblue3")+
  labs(
    title = "DWV Pathogen Abundance Across Different Host Plant Species",
    y = "Disease Abundance",
    x = "Host Plant",
    color = "Disease"
  )
q2_plot_dwv + coord_flip()


q2_anova <- manova(
  cbind(log10_Nosema_load,
    log10_DWV_load,
    log10_BQCV_load) ~ host_plant,
  data = bee_dat
)

summary(q2_anova)
```

### there isn't a statistically significant difference between different host plants and disease abundances, as the p-value was 0.5337
### I constructed the model like this because host plants are a catagorical variable, and this plot effectively shows disease abundance for each plant. 
### I removed sampling events from the data in case sampling events would have had an affect on pathogen abundance. If the sampling events were over a longer period of time, it could have potentially affected disease abundances



